---
title: "MATH 216 Homework 1"
author: "Bianca Gonzalez"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---
deleted cache = true; for unknown reasons
```{r, echo=FALSE, message=FALSE}

library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 
library(lubridate)

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>% 
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```

## Admistrative:

Please indicate

* Who you collaborated with:
Me, Myself, and Albert Kim. (office hours on Tuesday Oct 4th.)

* Roughly how much time you spent on this HW so far:
Likely three or four hours. I sped up as I practiced more throughout the week.

* The URL of the RPubs published URL [here](http://rpubs.com/biancaglez94/215369).

* What gave you the most trouble:
The month function. Understanding conceptually what was happening with the grouping function
and what variables to use. 

* Any comments you have:
I would like the homework to be worth more percentage points for our grades because they are where the learning happens. Or the class excercises being worth something. Incentives. 

## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
#proportion = nominator flights delayed >30/ denominator total flights

#flights delayed by thirty minutes (numerator)

  #Wanto to see the flights over months to rid of some noise in plot from days
  # when plotting carriers with delayed proportion. 
  flights <- mutate(flights, date2 = format(flights$date, "%m"))
  #find the number of flights delayed per carrier
  flights_delay <- flights %>% 
    filter(dep_delay >= 30) %>% 
    group_by(date2, dep_delay, carrier) %>% 
    tally() %>% 
    rename(num_flights_30 = n) %>% 
    arrange(desc(num_flights_30))
 # flights_delay
  

#this is the total number of flights per carrier
flights_tot <- flights %>%    # filter(carrier == "AA" || carrier == "DL")
  group_by(date2, carrier) %>% 
  tally() %>% 
  rename(num_flights = n)
#flights_tot

#join these datasets so we can mutuate and compute proportion
flights_prop <- inner_join(flights_tot, flights_delay, by="date2") %>% 
  mutate(proportion_flight_delayed = num_flights_30/num_flights)
flights_prop
#now to plot! 
#yaxis proportion. x axis time, so lets try months over this year. 
#add flight carrier?
ggplot(data=flights_prop, aes(x=date2, y=proportion_flight_delayed , col=carrier.y)) +
geom_line()

# I like this plot, but I don't like that there is so much noise, for so many carriers. Also, when looking at the data, I think there is room here for a statistical test to see what carriers are statistically signifcantly late. 

#These plots shows the proportion of the airlines flights that were late throughout the year. Prefer this plot b/c. 

ggplot(data=flights_prop, aes(x=date2, y=proportion_flight_delayed)) + facet_wrap(~carrier.y) +
geom_line()


```

## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?

year insufficient in providing the comfort of a passenger
so I'm using different proxies, including year to determine
what planes have more room. :

      -no.seats = less seats the more likely to have extra leg room!
      -no.eng = more engines, bigger plane!
      
by arranging in descending order of number of engines, and the number of seats, we find the 
planes that are the largest (number of engines) and have the least number of seats

Looks like CANADAIR LTD and AVIONS MARCEL DASSAULT have most legroom. 

```{r, echo=FALSE, fig.width=12, fig.height=6}

leg_room <- planes %>% 
  filter(!is.na(year)) %>%   
  group_by(no.eng, mfr, year) %>%  #to get size of planes by carriers (mfr)and years
  summarise(min_seats = min(no.seats)) %>%  #get the number of seats for size of planes
  arrange(desc(no.eng)) 
leg_room


#Let's just look at age of plane as per the assignment to find most legroom

legroom_yr <- planes %>% 
  filter(!is.na(year)) %>%   
  group_by(mfr, year) %>%  #to get size of planes by carriers (mfr)and years
  summarise(min_seats = min(no.seats)) %>%  #get the number of seats for size of planes
  arrange(year, min_seats) 
                          #question, can I arrange by two variables?? ie the two lowest variables                           followed by the next two lowest, and so on. 
legroom_yr
```
## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.
```{r, echo=TRUE, message=FALSE, fig.width=12, fig.height=6}

#lets first join airports and flights to find states SW flies to. 
sw_join <- full_join(airports, flights, by=c("iata"="dest"))

#dataset with only SouthWest airline flights and no NAs. 
sw_flights <- sw_join %>% 
  filter(!is.na(flight)) %>% 
  filter((carrier == "WN")) 
sw_flights

sw_tot <- sw_flights %>% 
  select(flight) %>% 
  tally() %>% 
  rename(tot_sw = n)
sw_tot

#find state where flight paths tend to fly to (so dont group by destination)
sw_fp <- sw_flights %>% 
  group_by(flight, state) %>% 
  tally() %>% 
  rename(flight_dest = n) %>% 
  arrange(desc(flight_dest))
sw_fp

#gives us the common flight destinations for flight paths.

#find states flights tend to fly to 
sw_states <- sw_flights %>% 
  select(state, iata) %>% 
  group_by(state, iata) %>% 
  tally() %>% 
  rename(state_count = n) %>% 
  arrange(desc(state_count))
sw_states


```

## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=TRUE, message=FALSE, fig.width=12, fig.height=6}
library(lubridate)

#month function. proportionally. carrier region total/ flight total
#carrier region total/ flight total. 

#sw_join was data made earlier, joiningg airoprts and flights dataset togehter. 
sw_join <- full_join(airports, flights, by=c("iata"="dest"))
region <- full_join(sw_join, states, by = "state")


July_Flight <- region %>% 
  filter(!is.na(carrier) & (!is.na(date)))  %>% 
  mutate(month = month(date)) %>% 
  filter(month == 7) 
 # filter(carrier == "AA")  Testing my proportion by checking to see if AA only flew to a                   Southern regions as computed by our proportion below. checks out!
                            
regionTot <- July_Flight %>% 
  group_by(carrier, region) %>% 
  tally() %>% 
  rename(region_fly_tot = n) %>%  
arrange(desc(region_fly_tot))     #looks like most flights are going to the South.                                             connecting flights perhaps?

flight_tot <- July_Flight %>% 
  group_by(carrier) %>% 
  tally() %>% 
  rename(tot_flights = n) %>% 
arrange(desc(tot_flights))       #alright now we have all flights by carriers! 
  
#to compute this proportion, let's join region flight totals and flights totals looks like the 
props <- left_join(regionTot, flight_tot, by="carrier") %>% 
  mutate(prop_region = (region_fly_tot/tot_flights)) 
 # arrange(carrier)

```
