---
title: "MATH 216 Homework 1"
author: "Bianca Gonzalez"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    smooth_scroll: false
---

```{r, echo=FALSE, message=FALSE}
library(knitr)
library(lubridate)
library(ggplot2)
library(dplyr)
# For read_csv() command, which is an improved version of base R's read.csv()
library(readr) 
library(lubridate)

# Load data sets. Note this assumes this .Rmd files is in the same directory as
# the .csv files.
flights <- read_csv("data/flights.csv") %>%
  mutate(date=as.Date(date))
weather <- read_csv("data/weather.csv") %>% 
  mutate(date=as.Date(date))
planes <- read_csv("data/planes.csv")
airports <- read_csv("data/airports.csv")
states <- read_csv("data/states.csv")
```

## Admistrative:

Please indicate

* Who you collaborated with:
Me, Myself, and Albert Kim. (office hours on Tuesday Oct 4th.)

* Roughly how much time you spent on this HW so far:
Likely three or four hours. I sped up as I practiced more throughout the week.

* The URL of the RPubs published URL [here](http://rpubs.com/biancaglez94/215369).

* What gave you the most trouble:
The month function. Understanding conceptually what was happening with the grouping function
and what variables to use. 

* Any comments you have:
I would like the homework to be worth more percentage points for our grades because they are where the learning happens. Or the class excercises being worth something. Incentives. 

## Question 1:

Plot a "time series" of the proportion of flights that were delayed by > 30 minutes on each day.  i.e.
 
* the x-axis should be some notion of time
* the y-axis should be the proportion.

Using this plot, indicate describe the
[seasonality](https://en.wikipedia.org/wiki/Seasonality) of when delays over 30
minutes tend to occur.

```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
#proportion = nominator flights delayed >30/ denominator total flights

#flights delayed by thirty minutes (numerator)

  #Wanto to see the flights over months to rid of some noise in plot from days
  # when plotting carriers with delayed proportion. 
  flights <- mutate(flights, date2 = format(flights$date, "%m"))
  #find the number of flights delayed per carrier
  flights_delay <- flights %>% 
    filter(dep_delay >= 30) %>% 
    group_by(date2) %>% 
    tally() %>% 
    rename(num_flights_30 = n) %>% 
    arrange(desc(num_flights_30))
  knitr::kable(flights_delay)
 
#this is the total number of flights
flights_tot <- flights %>%  
  group_by(date2) %>% 
  tally() %>% 
  rename(num_flights = n)
#flights_tot

#join these datasets so we can mutuate and compute proportion
flights_prop <- inner_join(flights_tot, flights_delay, by="date2") %>% 
  mutate(proportion_flight_delayed = num_flights_30/num_flights)
knitr::kable(flights_prop)
#now to plot! 
#yaxis proportion. x axis time, so lets try months over this year. 

#1. what am I plotting as line? why not the proportion value?
flights_prop %>% 
  mutate(date2=as.numeric(date2)) %>% 
ggplot(aes(x=date2, y=proportion_flight_delayed)) +
geom_line() + 
  scale_x_continuous(breaks=c(1:12)) +
         labs(x = "Months", y = "Proportion of flights delayed", 
              title = "Proportion of Flights delayed more than 30 minutes")

```


```{r, echo=FALSE, message=FALSE, fig.width=12, fig.height=6}
#These plots shows the proportion of the airlines flights that were late throughout the year. 
#ggplot(data=flights_prop, aes(x=date2, y=proportion_flight_delayed)) + facet_wrap(~carrier.y) +
#geom_line()

```

## Question 2:

Some people prefer flying on older planes.  Even though they aren't as nice,
they tend to have more room.  Which airlines should these people favor?


```{r, echo=FALSE, fig.width=12, fig.height=6}
#Let's just look at age of plane as per the assignment to find most legroom
#need to find not the airplane, but airline... so join with dataset flights to get carrier
#also need to find mean number of years for each carrier. 
legroom_yr <- inner_join(planes, flights, by= "plane")

legroom_yr %>% 
  filter(!is.na(year)) %>%   
  group_by(carrier) %>%  #to get size of planes by carriers (mfr)and years
  summarise(mean_years = mean(year))   #get the number of seats for size of planes
#knitr::kable(legroom_yr)


```

## Question 3:

* What states did Southwest Airlines' **flight paths** tend to fly to?
* What states did Southwest Airlines' **flights** tend to fly to?

For example, Southwest Airlines Flight 60 to Dallas consists of a single flight
path, but since it flew 299 times in 2013, it would be counted as 299 flights.
```{r, echo=TRUE, message=FALSE, fig.width=12, fig.height=6}

#lets first join airports and flights to find states SW flies to. 
sw_join <- full_join(airports, flights, by=c("iata"="dest"))

#dataset with only SouthWest airline flights and no NAs. 
sw_flights <- sw_join %>% 
  filter(!is.na(flight)) %>% 
  filter((carrier == "WN")) 

sw_tot <- sw_flights %>% 
  select(flight) %>% 
  tally() %>% 
  rename(tot_sw = n)

#find state where flight paths tend to fly to (so dont group by destination)
sw_fp <- sw_flights %>% 
  group_by(flight, state) %>% 
  tally() %>% 
  rename(flight_dest = n)
sw_fp
   # head(10) %>%
  #  format(big.mark = ",") %>%
   # knitr::kable(sw_fp)

#gives us the common flight destinations for flight paths.
#find states flights tend to fly to 
sw_states <- sw_flights %>% 
  select(state, iata) %>% 
  group_by(state, iata) %>% 
  tally() %>% 
  rename(state_count = n) %>% 
  head(10) %>%
    format(big.mark = ",") %>%
    knitr::kable("markdown", align = c('l', 'r'))


```

## Question 4:

I want to know proportionately what regions (NE, south, west, midwest) each 
carrier flies to/from Houston in the month of July.  Consider the `month()`
function from the `lubridate` package.

```{r, echo=TRUE, message=FALSE, fig.width=12, fig.height=6}

#month function. proportionally. carrier region total/ flight total
#carrier region total/ flight total. 

#sw_join was data made earlier, joiningg airoprts and flights dataset togehter. 
sw_join <- full_join(airports, flights, by=c("iata"="dest"))
region <- full_join(sw_join, states, by = "state")


July_Flight <- region %>% 
  filter(!is.na(carrier) & (!is.na(date)))  %>% 
  mutate(month = month(date)) %>% 
  filter(month == 7) 
                            
regionTot <- July_Flight %>% 
  group_by(region, carrier) %>% 
  tally() %>% 
  rename(region_fly_tot = n) 

flight_tot <- July_Flight %>% 
  group_by(region) %>%       
  tally() %>% 
  rename(tot_flights = n) 
```

```{r, echo=TRUE, message=FALSE, fig.width=12, fig.height=6}
#to compute this proportion, let's join region flight totals and flights totals 
props <- left_join(regionTot, flight_tot, by="region") %>% 
  mutate(prop_region = (region_fly_tot/tot_flights)) 

#mutate(props, prop_region=as.numeric(prop_region)) 

ggplot(data = props, aes(x=carrier, y=prop_region, col = region)) +
geom_point() +
         labs(x = "Airline Carrier", y = "Proportion", 
              title = "Proportionately what regions each 
airline carrier flies to in July")

```
